paper figures

### 2025-10-22 paper_figures.Rmd
re-order source data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, cache.lazy=FALSE, message=FALSE, warning=FALSE,dev=c('pdf'))
library(ggpubr)
library(stats)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(caret)
library(scales)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(ggrepel)
library(car)
library(glmnet)
library(pROC)
library(readxl)
library(variancePartition)
library(RColorBrewer)
library(gtools)
library(grid)
library(gtools)
library(lme4)
library(dendextend)
```

@@

```{r Fig1D, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
# simulation data for Fig. 1
synthetic.data <- read.csv('data/synthetic_data.csv',header=1, row.names=1)
ggplot(synthetic.data %>% 
         dplyr::mutate(nreads=factor(nreads, levels=c('10k','20k','50k','100k'))), 
       aes(x=nclones, y=clusters_raw, color=nreads)) + 
  geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
  geom_point(size=.5) + 
  geom_smooth(method='lm', alpha=.2, size=.5, se=FALSE) +
  scale_x_log10() + 
  scale_y_log10() +
  my_theme() +
  scale_color_brewer(palette='Set2') +
  stat_cor(method = "pearson", aes(label=..rr.label..), size=2, show.legend=FALSE) +
  theme(legend.position=c(.85,.35),
        legend.background = element_rect(fill=NA)) +
  labs(x='# Clones', y='# Filtered clusters', color='# Reads')
```

@@

```{r Fig1E, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
# cell number experiment for  Fig. 1
cellswitch.data <- read.csv('data/cellswitch_data.csv',header=1, row.names=1)

@@

```{r Fig1F, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
meta_clustering <- read.csv('data/cellswitch_meta_clustering.csv', header=1, row.names=1)

ggplot(meta_clustering %>% 
         dplyr::mutate(ncells=factor(ncells, levels=c('50k','100k')),
                       Donor=ifelse(Donor=='mixed','mixed',paste0(Donor,'-',Donor)),
                       mixing=gsub('-donor','',mixing)), 
       aes(x=ncells,y=100*frac_clusters, fill=mixing)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA) +
  geom_jitter(aes(color=Donor, group=mixing), position=position_jitterdodge(dodge.width = .8, jitter.width=.05), size=1) +
  my_theme() +
  #theme(legend.position='bottom') +
  #coord_cartesian(ylim=c(15,40)) +
  scale_fill_manual(values=c('gray','white'), guide=FALSE) +
  scale_color_manual(values=setNames(c(brewer.pal(3,'Set1'),'gray20'),c('A-A','B-B','C-C','mixed')),na.value='gray20') +
  labs(x='# cells',y='% shared clusters in pair', fill='pair', color='Donors')
```

@@

```{r Fig1G, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
# bulk VDJ results for Fig. 1
VDJ_results <- read.csv('data/VDJ_results.csv',header=1, row.names=1)

@@

```{r FigS1X1, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
vdj_df %>% 
  dplyr::mutate(ncells=factor(gsub('000$','k',ncells), levels=c('50k','100k')),
                Donor=gsub('Donor ','',Donor)) %>%
  dplyr::filter(metric=='cluster_gini_raw') %>%
  dplyr::group_by(ncells, Donor, method) %>%
  dplyr::summarise(mean_gini=mean(value),
                   se_gini=sd(value)/sqrt(length(value))) %>%
  pivot_wider(id_cols=c(ncells,Donor),names_from=method,values_from=c('mean_gini','se_gini')) %>%
  ggplot(aes(x=mean_gini_VDJ, y=mean_gini_SWIBRID,
             xmin=mean_gini_VDJ-se_gini_VDJ,
             xmax=mean_gini_VDJ+se_gini_VDJ,
             ymin=mean_gini_SWIBRID-se_gini_SWIBRID,
             ymax=mean_gini_SWIBRID+se_gini_SWIBRID)) +
  geom_point(aes(color=Donor,shape=ncells),size=1) +
  geom_errorbar(linewidth=.25) +
  geom_errorbarh(linewidth=.25) +
  #scale_x_continuous(breaks=c(.6,.65)) +
  my_theme() +
  scale_color_brewer(palette='Set1') +
  theme(strip.background=element_blank()) +
  labs(y='Gini coefficient SWIBRID',x='Gini coefficient VDJ',shape='# cells')
```

```{r FigS1X2, fig.width=2 * fig_scale, fig.height=1.4 * fig_scale}
library(ggpmisc)
PCR.data <- read.csv('data/PCR_data.csv',header=1,row.names=1)
ggplot(PCR.data, aes(x=as.integer(ncycles), y=log10(reads))) + 
  geom_point(aes(color=donor)) + 
  geom_smooth(method='lm', color='black', linewidth=.5) +
  geom_abline(slope=log10(2), intercept=-1.5, linewidth=.5, linetype='dashed') +
  my_theme() + 
  scale_color_brewer(palette='Set2') +
  scale_y_continuous(breaks = log10(c(1, 10, 100, 1000,10000, 100000)),
                     label = c(1, 10, 100, 1000, 10000, 100000)) +
  labs(x='# of cycles',y='# of reads')
```

@@

# HD data for Fig. 2
HD.data <- read.csv('data/HD_data.csv', header=1, row.names=1)

@@

```{r Fig2D, fig.width=1.5 * fig_scale,fig.height=1.75 * fig_scale}
donor_cutoff <- .25
batch_cutoff <- .35
sequencing_cutoff <- .2
selected.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('Diversity_raw','Context_isotype','library')), 
                !(col %in% c('log10_nreads','mean_GC')), 
                Donor >= donor_cutoff, 
                Sequencing < sequencing_cutoff, 
                Batch < batch_cutoff) %>%
  dplyr::arrange(category) %>%
  dplyr::pull(col)

@@

```{r FigS2J_1, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable,
                                levels=rev(c('Donor','Batch','Sequencing','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('Diversity','Diversity_raw'))

@@

```{r FigS2J_2, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable,
                                levels=rev(c('Donor','Batch','Sequencing','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('Context','Context_isotype'))

@@

```{r FigS2J_3, fig.width=1.4 * fig_scale,fig.height=1.7 * fig_scale}
HD.data.br <- read.csv('data/HD_data_by_reads.csv', header=1, row.names=1)

@@

```{r Fig2C, fig.width=4.25 * fig_scale, fig.height=1.5 * fig_scale}
cols.here.human <- c('pct_templated_inserts',
                     'untemplated_inserts',
                     'pct_sequential_switch',
                     'pct_intraswitch_deletion',
                     'cluster_gini',
                     'pct_clusters_SG1')

@@

```{r Fig2F, fig.width=4. * fig_scale, fig.height=1.75 * fig_scale}
mouse.data <- read.csv('data/mouse_data.csv',header=1, row.names=1)  %>%
  tibble::column_to_rownames('Sample')
vincendeau.data <- read.csv('data/vincendeau_data.csv', header=1, row.names=1)  %>%
  tibble::column_to_rownames('Sample')

@@

```{r FigS2K, fig.width=1.5 * fig_scale, fig.height=1.75 * fig_scale}
cols.here.mouse <- c('read_length_SG_mean')
cohort.data <- rbind(mouse.data %>%
                       dplyr::mutate(species='mouse tissue',
                                     Cohort=material) %>%
                       dplyr::select(species, Cohort, cols.here.mouse),
                     vincendeau.data %>%
                       dplyr::mutate(species='Vincendeau',
                                     Cohort=Genotype) %>%
                       dplyr::select(species, Cohort, cols.here.mouse)) %>%
  dplyr::mutate(Cohort=factor(Cohort, levels=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'))) %>%
  gather(variable, value, cols.here.mouse) %>%
  dplyr::select(variable, value, species, Cohort) %>%
  dplyr::left_join(columns, by=c('variable'='col')) %>%
  dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural'))

@@

plots <- lapply(comparisons, function(x) 
  tmp %>%
    dplyr::mutate_(mean1=paste0('mean_',x[[1]]),
                   mean2=paste0('mean_',x[[2]]),
                   se1=paste0('se_',x[[1]]),
                   se2=paste0('se_',x[[2]])) %>%
    dplyr::mutate(tstat=abs(mean2-mean1)/(sqrt(se2**2+se1**2)),
                  pval=dt(tstat,4)) %>%
    dplyr::mutate(padj=p.adjust(pval, method='BH'),
                  sig=ifelse(padj < .05, 'sig','ns')) %>%
    ggplot(aes(x=mean1,y=mean2)) +
    geom_abline(slope=1,intercept=0,color='gray',linewidth=.5) +
    geom_point(aes(color=variable, shape=variable, size=sig, alpha=sig)) + 
    stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE) +
    geom_errorbar(aes(ymin=mean2-se2,
                      ymax=mean2+se2,
                      color=variable,
                      alpha=sig),
                  size=.25) +
    geom_errorbarh(aes(xmin=mean1-se1,
                       xmax=mean1+se1,
                       color=variable,
                       alpha=sig),
                   size=.25) +
    my_theme() +
    labs(color='',shape='') +
    scale_color_manual(values=cols_shapes %>% 
                         dplyr::pull(var_col,variable),
                       labels=cols_shapes %>%
                         dplyr::pull(variable)) +
    scale_shape_manual(values=cols_shapes %>%
                         dplyr::pull(var_ind,variable),
                       labels=cols_shapes %>%
                         dplyr::pull(variable)) +
    scale_size_manual(values=c('sig'=2,'ns'=.5),guide='none') +
    scale_alpha_manual(values=c('sig'=1,'ns'=.25),guide='none') +
    guides(color=guide_legend(ncol=5),
           shape=guide_legend(ncol=5)) +
    labs(x=x[[1]],y=x[[2]],fill='',color='',size=''))

plot_grid(plot_grid(plotlist=lapply(plots, function(x) x + theme(legend.position='none')), 
                    ncol=4, align='vh'),
          get_legend(plots[[1]] + theme(legend.margin=margin(0,0,0,0,'mm'))), ncol=1, rel_heights=c(.65,.35))
```

<!-- the PCR cycles experiment -->

<!-- ```{r PCR_cycles} -->
<!-- PCR.data <- read.csv('../human_samples/20250703_PCR_cycles/output/summary/20250703_PCR_cycles_stats.csv',header=1,row.names=1) %>% -->
<!--   tibble::rownames_to_column('sample') %>% -->
<!--   dplyr::mutate(donor=gsub('[0-9]*_(P[0-9])x([0-9]*)','\\1',sample), -->
<!--                 ncycles=gsub('[0-9]*_(P[0-9])x([0-9]*)','\\2',sample))  -->

<!-- PCR.X <- PCR.data[,c('sample',selected.cols)] %>% -->
<!--   dplyr::mutate_if(is.numeric,scale) %>% -->
<!--   tibble::column_to_rownames('sample') -->

<!-- PCR.meta <- PCR.data[,c('ncycles','donor')] -->

<!-- cols.here <- c('reads','clusters_raw','PCR_bias_length','PCR_bias_GC') -->
<!-- PCR.data %>% -->
<!--   gather(variable, value, cols.here) %>% -->
<!--   dplyr::select(variable, value, ncycles, donor) %>% -->
<!--   ggplot(aes(x=ncycles, y=value, fill=ncycles)) +  -->
<!--   geom_boxplot(position=position_dodge(width=.75), width=.7, outlier.shape=NA, lwd=.25) +  -->
<!--   geom_point(aes(color=donor), position=position_jitterdodge(jitter.width=.2, dodge.width=.75, jitter.height = 0), size=.5) +  -->
<!--   my_theme() + -->
<!--   coord_cartesian(clip='off') + -->
<!--   facet_wrap(~variable, switch='y', scales='free_y') + -->
<!--   theme(axis.text.x=element_text(angle=45, hjust=1, vjust=1), -->
<!--         legend.position='none') + -->
<!--   labs(x='',y='', fill='')  -->

<!-- take <- (rowSums(is.na(PCR.X))==0) &  (PCR.meta$ncycles %in% c('15','20','25')) -->
<!-- pca <- prcomp(PCR.X[take,]) -->
<!-- pcadat <- cbind(pca$x,PCR.meta[take,]) -->

<!-- ggplot(pcadat,aes(x=PC1,y=PC2)) + -->
<!--   geom_point(aes(color=donor, shape=ncycles),size=3) + -->
<!--   my_theme() + -->
<!--   guides(color=guide_legend(override.aes = list(shape=15,size=2), ncol=3)) + -->
<!--   theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2, -->
<!--         legend.key.size=unit(1.5,'mm'), -->
<!--         legend.key.spacing.y=unit(0,'mm'), -->
<!--         legend.position='bottom', -->
<!--         legend.box='horizontal') + -->
<!--   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'), -->
<!--        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'), -->
<!--        color='',shape='',fill='') -->
<!-- ``` -->

<!-- ```{r cell_switch} -->
<!-- cellswitch_sample_sheet <- read.csv('../sodar/2020_CellSwitch/a_2020_CellSwitch.txt', -->
<!--                             sep='\t',header=1) %>% -->
<!--   dplyr::left_join(read.csv('../sodar/2020_CellSwitch/s_2020_CellSwitch.txt',sep='\t',header=1), -->
<!--                    by='Sample.Name') %>% -->
<!--   dplyr::mutate(Donor=as.character(Source.Name), -->
<!--                 sample=Sample.Name, -->
<!--                 ncells=Parameter.Value.Cell.Number., -->
<!--                 Batch=as.character(Parameter.Value.Batch.), -->
<!--                 QC=as.character(Parameter.Value.QC.)) %>% -->
<!--   dplyr::select(Donor,sample,ncells,Batch,QC) %>% -->
<!--   dplyr::filter(Donor %in% c('21084','21085','21086')) -->

<!-- cellswitch.data <- read.csv('../human_samples/20240307_hg38_results/output/summary/20240307_human_samples_stats.csv', -->
<!--                           header=1,row.names=1) %>% -->
<!--   tibble::rownames_to_column('sample') %>% -->
<!--   dplyr::inner_join(cellswitch_sample_sheet, by='sample') %>% -->
<!--   dplyr::filter(reads >= 500, QC=='PASS') %>% -->
<!--   dplyr::mutate_all(~replace_na(.,0)) %>% -->
<!--   dplyr::mutate(Donor=plyr::revalue(Donor, c('21084'='A', -->
<!--                                              '21085'='B', -->
<!--                                              '21086'='C'))) %>% -->
<!--   dplyr::select(Donor,ncells,Batch,sample,selected.cols) -->

<!-- cellswitch.X <- cellswitch.data[,c('sample',selected.cols)] %>% -->
<!--   dplyr::mutate_if(is.numeric,scale) %>% -->
<!--   tibble::column_to_rownames('sample') -->

<!-- cellswitch.meta <- cellswitch.data[,c('Donor','ncells')] -->

<!-- take <- (rowSums(is.na(cellswitch.X))==0) #& (cellswitch.meta$ncells > 10000) -->
<!-- pca <- prcomp(cellswitch.X[take,]) -->
<!-- pcadat <- cbind(pca$x,cellswitch.meta[take,]) -->

<!-- ggplot(pcadat,aes(x=PC1,y=PC2)) + -->
<!--   geom_point(aes(fill=ncells),size=3,pch=21) + -->
<!--   geom_line(aes(group=Donor,color=Donor)) + -->
<!--   my_theme() + -->
<!--   theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2, -->
<!--         legend.key.size=unit(1.5,'mm'), -->
<!--         legend.key.spacing.y=unit(0,'mm'), -->
<!--         legend.position='right') + -->
<!--   scale_fill_distiller(palette='Spectral',limits=c(100,200000), trans='log', breaks=c(1000,10000,100000)) + -->
<!--   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'), -->
<!--        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'), -->
<!--        color='Donor',fill='# cells') -->

<!-- take <- (rowSums(is.na(cellswitch.X))==0) & (cellswitch.meta$ncells > 10000) -->
<!-- pca <- prcomp(cellswitch.X[take,]) -->
<!-- pcadat <- cbind(pca$x,cellswitch.meta[take,]) -->

<!-- ggplot(pcadat,aes(x=PC1,y=PC2)) + -->
<!--   geom_point(aes(fill=ncells,color=Donor),size=3,pch=21) + -->
<!--   geom_line(aes(group=Donor,color=Donor)) + -->
<!--   my_theme() + -->
<!--   theme(aspect.ratio=pca$sdev[2]**2/pca$sdev[1]**2, -->
<!--         legend.key.size=unit(1.5,'mm'), -->
<!--         legend.key.spacing.y=unit(0,'mm'), -->
<!--         legend.position='right') + -->
<!--   scale_fill_distiller(palette='Spectral',limits=c(100,200000), trans='log', breaks=c(1000,10000,100000)) + -->
<!--   labs(x=paste0('PC 1 (',100*signif(pca$sdev[1]**2/sum(pca$sdev**2),3),'%)'), -->
<!--        y=paste0('PC 2 (',100*signif(pca$sdev[2]**2/sum(pca$sdev**2),3),'%)'), -->
<!--        color='Donor',fill='# cells') -->
<!-- ``` -->

# Figure 3 + supplement (CH12)

```{r FigS3A, fig.width=5.5 * fig_scale, fig.height=4.5 * fig_scale}
HTGTS.data <- read.csv("data/HTGTS_data.csv", header=1, row.names=1) %>%
  tibble::column_to_rownames('Sample')

@@

plots <- lapply(comparisons, function(x) 
  tmp %>%
    dplyr::mutate_(mean1=paste0('mean_',x[[1]],'_',x[[2]]),
                   mean2=paste0('mean_',x[[1]],'_',x[[3]]),
                   se1=paste0('se_',x[[1]],'_',x[[2]]),
                   se2=paste0('se_',x[[1]],'_',x[[3]])) %>%
    dplyr::mutate(tstat=abs(mean2-mean1)/(sqrt(se2**2+se1**2)),
                  pval=dt(tstat,4),
                  padj=p.adjust(pval, method='BH'),
                  sig=ifelse(padj < .05, 'sig','ns')) %>%
    ggplot(aes(x=mean1,y=mean2)) +
    geom_abline(slope=1,intercept=0,color='gray',linewidth=.5) +
    geom_point(aes(color=variable,shape=variable,size=sig, alpha=sig)) + 
    stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE, label.y.npc = 'bottom') +
    geom_errorbar(aes(ymin=mean2-se2,
                      ymax=mean2+se2,
                      color=variable,
                      alpha=sig),
                  size=.25) +
    geom_errorbarh(aes(xmin=mean1-se1,
                       xmax=mean1+se1,
                       color=variable,
                       alpha=sig),
                   size=.25) +
    my_theme() +
    labs(color='',shape='') +
    scale_color_manual(values=cols_shapes %>% 
                         dplyr::pull(var_col,variable),
                       labels=cols_shapes %>%
                         dplyr::pull(variable)) +
    scale_shape_manual(values=cols_shapes %>%
                         dplyr::pull(var_ind,variable),
                       labels=cols_shapes %>%
                         dplyr::pull(variable)) +
    scale_size_manual(values=c('sig'=2,'ns'=.5),guide='none') +
    scale_alpha_manual(values=c('sig'=1,'ns'=.2),guide='none') +
    guides(color=guide_legend(ncol=4),
           shape=guide_legend(ncol=4)) +
    labs(x=x[[2]],y=x[[3]],title=x[[1]],fill='',color='',size=''))

@@

```{r FigS3D, fig.width=3.25 * fig_scale, fig.height=1.25 * fig_scale}
read.csv('data/HTGTS_homology.csv',header=1, row.names=1) %>%
  dplyr::mutate(Genotype=factor(Genotype, levels=c('WT','Atm','Lig4','Trp53bp1')),
                celltype=gsub('spleen','Spleen',celltype)) %>%
  dplyr::filter(xor(isotype=='Sa', celltype=='Spleen')) %>%
  dplyr::group_by( celltype, Genotype, isotype, n_homology) %>%
  dplyr::summarise(mean_frac=mean(frac),
                   se_frac=sd(frac)/sqrt(n())) %>%
  ggplot(aes(x=n_homology, y=mean_frac, color=Genotype)) +
  geom_line(aes(group=Genotype)) +
  geom_errorbar(aes(ymin=mean_frac-se_frac,ymax=mean_frac+se_frac),width=0) +
  ggh4x::facet_nested(. ~ celltype  + isotype , scales='free', space='free_x', nest_line = TRUE, switch='y') +
  my_theme() +
  scale_color_manual(values=genotype_colors) +
  labs(color='', y='Fraction junctions') +
  coord_cartesian(clip='off') 
```

@@

```{r Fig3B, fig.width=3.5 * fig_scale, fig.height=1.75 * fig_scale}
CH12.data <- read.csv('data/CH12_data.csv',header=1, row.names=1) %>%
  tibble::column_to_rownames('Sample')
CH12.data.noSg <- read.csv('data/CH12_data_noSg.csv', header=1, row.names=1) %>%
  tibble::column_to_rownames('Sample')
mouse.data.noSg <- read.csv('data/mouse_data_noSg.csv', header=1, row.names=1)

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, cache.lazy=FALSE, message=FALSE, warning=FALSE,dev=c('pdf'))
library(ggpubr)
library(stats)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(caret)
library(scales)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(ggrepel)
library(car)
library(glmnet)
library(pROC)
library(readxl)
library(variancePartition)
library(RColorBrewer)
library(gtools)
library(grid)
library(gtools)
library(lme4)
library(dendextend)
library(openxlsx)
```

@@

```{r Fig1D, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
# simulation data for Fig. 1
#synthetic.data <- read.csv('data/synthetic_data.csv',header=1, row.names=1)
synthetic.data <- read.xlsx('data/source_data.xlsx',sheet='synthetic data')
ggplot(synthetic.data %>% 
         dplyr::mutate(nreads=factor(nreads, levels=c('10k','20k','50k','100k'))), 
       aes(x=nclones, y=clusters_raw, color=nreads)) + 
  geom_abline(slope=1,intercept=0, size=.5, color='gray80') +
  geom_point(size=.5) + 
  geom_smooth(method='lm', alpha=.2, size=.5, se=FALSE) +
  scale_x_log10() + 
  scale_y_log10() +
  my_theme() +
  scale_color_brewer(palette='Set2') +
  stat_cor(method = "pearson", aes(label=..rr.label..), size=2, show.legend=FALSE) +
  theme(legend.position=c(.85,.35),
        legend.background = element_rect(fill=NA)) +
  labs(x='# Clones', y='# Filtered clusters', color='# Reads')
```

@@

```{r Fig1E, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
# cell number experiment for  Fig. 1
cellswitch.data <- read.xlsx('data/source_data.xlsx',sheet='cell numbers')

@@

```{r Fig1F, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
meta_clustering <- read.xlsx('data/source_data.xlsx', sheet='meta-clustering')

ggplot(meta_clustering %>% 
         dplyr::mutate(ncells=factor(ncells, levels=c('50k','100k')),
                       Donor=ifelse(Donor=='mixed','mixed',paste0(Donor,'-',Donor)),
                       mixing=gsub('-donor','',mixing)), 
       aes(x=ncells,y=100*frac_clusters, fill=mixing)) + 
  geom_boxplot(position=position_dodge(width=.8), outlier.shape=NA) +
  geom_jitter(aes(color=Donor, group=mixing), position=position_jitterdodge(dodge.width = .8, jitter.width=.05), size=1) +
  my_theme() +
  scale_fill_manual(values=c('gray','white'), guide=FALSE) +
  scale_color_manual(values=setNames(c(brewer.pal(3,'Set1'),'gray20'),c('A-A','B-B','C-C','mixed')),na.value='gray20') +
  labs(x='# cells',y='% shared clusters in pair', fill='pair', color='Donors')
```

@@

```{r Fig1G, fig.width=1.4 * fig_scale, fig.height=1.4 * fig_scale}
# bulk VDJ results for Fig. 1
VDJ_results <- read.xlsx('data/source_data.xlsx',sheet='VDJ data')

@@

@@

# HD data for Fig. 2
HD.data <- read.xlsx('data/source_data.xlsx', sheet='HD data')

@@

```{r Fig2C, fig.width=1.5 * fig_scale,fig.height=1.75 * fig_scale}
donor_cutoff <- .25
batch_cutoff <- .35
sequencing_cutoff <- .2
selected.cols <- HD.var %>%
  dplyr::left_join(columns, by='col') %>%
  spread(variable, variance) %>%
  dplyr::filter(!(category %in% c('Diversity_raw','Context_isotype','library')), 
                !(col %in% c('log10_nreads','mean_GC')), 
                Donor >= donor_cutoff, 
                Sequencing < sequencing_cutoff, 
                Batch < batch_cutoff) %>%
  dplyr::arrange(category) %>%
  dplyr::pull(col)

@@

```{r FigS2I_1, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable,
                                levels=rev(c('Donor','Batch','Sequencing','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('Diversity','Diversity_raw'))

@@

```{r FigS2I_2, fig.width=1.4 * fig_scale, fig.height=1.7 * fig_scale}
tmp <- HD.var %>%  
  dplyr::left_join(columns, by='col') %>%
  dplyr::mutate(variable=factor(variable,
                                levels=rev(c('Donor','Batch','Sequencing','Residuals')))) %>%
  dplyr::filter(variable!='Residuals', category %in% c('Context','Context_isotype'))

@@

```{r FigS2I_3, fig.width=1.4 * fig_scale,fig.height=1.7 * fig_scale}
HD.data.br <- read.xlsx('data/source_data.xlsx', sheet='HD data by reads')

@@

```{r Fig2D, fig.width=4.25 * fig_scale, fig.height=1.5 * fig_scale}
cols.here.human <- c('pct_templated_inserts',
                     'untemplated_inserts',
                     'pct_sequential_switch',
                     'pct_intraswitch_deletion',
                     'cluster_gini',
                     'pct_clusters_SG1')

@@

```{r Fig2F, fig.width=4. * fig_scale, fig.height=1.75 * fig_scale}
mouse.data <- read.xlsx('data/source_data.xlsx', sheet='mouse tissues') %>%
  tibble::column_to_rownames('Sample')
vincendeau.data <- read.xlsx('data/source_data.xlsx', sheet='Vincendeau data') %>%
  tibble::column_to_rownames('Sample')

@@

```{r FigS2J, fig.width=1.5 * fig_scale, fig.height=1.75 * fig_scale}
cols.here.mouse <- c('read_length_SG_mean')
cohort.data <- rbind(mouse.data %>%
                       dplyr::mutate(species='mouse tissue',
                                     Cohort=material) %>%
                       dplyr::select(species, Cohort, cols.here.mouse),
                     vincendeau.data %>%
                       dplyr::mutate(species='Vincendeau',
                                     Cohort=Genotype) %>%
                       dplyr::select(species, Cohort, cols.here.mouse)) %>%
  dplyr::mutate(Cohort=factor(Cohort, levels=c('lymph node','splenocytes','WT','Xlf','Shld1','Shld1_Xlf'))) %>%
  gather(variable, value, cols.here.mouse) %>%
  dplyr::select(variable, value, species, Cohort) %>%
  dplyr::left_join(columns, by=c('variable'='col')) %>%
  dplyr::filter(category %in% c('Diversity','Isotypes','Breaks','Context','Structural'))

@@

plots <- lapply(comparisons, function(x) 
  tmp %>%
    dplyr::mutate_(mean1=paste0('mean_',x[[1]]),
                   mean2=paste0('mean_',x[[2]]),
                   se1=paste0('se_',x[[1]]),
                   se2=paste0('se_',x[[2]])) %>%
    dplyr::mutate(tstat=abs(mean2-mean1)/(sqrt(se2**2+se1**2)),
                  pval=2*pt(-abs(tstat),4)) %>%
    dplyr::mutate(padj=p.adjust(pval, method='BH'),
                  sig=ifelse(padj < .05, 'sig','ns')) %>%
    ggplot(aes(x=mean1,y=mean2)) +
    geom_abline(slope=1,intercept=0,color='gray',linewidth=.5) +
    geom_point(aes(color=variable, shape=variable, size=sig, alpha=sig)) + 
    stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE) +
    geom_errorbar(aes(ymin=mean2-se2,
                      ymax=mean2+se2,
                      color=variable,
                      alpha=sig),
                  size=.25) +
    geom_errorbarh(aes(xmin=mean1-se1,
                       xmax=mean1+se1,
                       color=variable,
                       alpha=sig),
                   size=.25) +
    my_theme() +
    labs(color='',shape='') +
    scale_color_manual(values=cols_shapes %>% 
                         dplyr::pull(var_col,variable),
                       labels=cols_shapes %>%
                         dplyr::pull(variable)) +
    scale_shape_manual(values=cols_shapes %>%
                         dplyr::pull(var_ind,variable),
                       labels=cols_shapes %>%
                         dplyr::pull(variable)) +
    scale_size_manual(values=c('sig'=2,'ns'=.5),guide='none') +
    scale_alpha_manual(values=c('sig'=1,'ns'=.25),guide='none') +
    guides(color=guide_legend(ncol=5),
           shape=guide_legend(ncol=5)) +
    labs(x=x[[1]],y=x[[2]],fill='',color='',size=''))

plot_grid(plot_grid(plotlist=lapply(plots, function(x) x + theme(legend.position='none')), 
                    ncol=4, align='vh'),
          get_legend(plots[[1]] + theme(legend.margin=margin(0,0,0,0,'mm'))), ncol=1, rel_heights=c(.65,.35))
```
# Figure 3 + supplement (CH12)

```{r FigS3A, fig.width=5.5 * fig_scale, fig.height=4.5 * fig_scale}
HTGTS.data <- read.xlsx("data/source_data.xlsx", sheet='HTGTS data') %>%
  tibble::column_to_rownames('Sample')

@@

plots <- lapply(comparisons, function(x) 
  tmp %>%
    dplyr::mutate_(mean1=paste0('mean_',x[[1]],'_',x[[2]]),
                   mean2=paste0('mean_',x[[1]],'_',x[[3]]),
                   se1=paste0('se_',x[[1]],'_',x[[2]]),
                   se2=paste0('se_',x[[1]],'_',x[[3]])) %>%
    dplyr::mutate(tstat=abs(mean2-mean1)/(sqrt(se2**2+se1**2)),
                  pval=2*pt(-abs(tstat),4),
                  padj=p.adjust(pval, method='BH'),
                  sig=ifelse(padj < .05, 'sig','ns')) %>%
    ggplot(aes(x=mean1,y=mean2)) +
    geom_abline(slope=1,intercept=0,color='gray',linewidth=.5) +
    geom_point(aes(color=variable,shape=variable,size=sig, alpha=sig)) + 
    stat_cor(method = "pearson", aes(label=..r.label..), size=2.5, show.legend=FALSE, label.y.npc = 'bottom') +
    geom_errorbar(aes(ymin=mean2-se2,
                      ymax=mean2+se2,
                      color=variable,
                      alpha=sig),
                  size=.25) +
    geom_errorbarh(aes(xmin=mean1-se1,
                       xmax=mean1+se1,
                       color=variable,
                       alpha=sig),
                   size=.25) +
    my_theme() +
    labs(color='',shape='') +
    scale_color_manual(values=cols_shapes %>% 
                         dplyr::pull(var_col,variable),
                       labels=cols_shapes %>%
                         dplyr::pull(variable)) +
    scale_shape_manual(values=cols_shapes %>%
                         dplyr::pull(var_ind,variable),
                       labels=cols_shapes %>%
                         dplyr::pull(variable)) +
    scale_size_manual(values=c('sig'=2,'ns'=.5),guide='none') +
    scale_alpha_manual(values=c('sig'=1,'ns'=.2),guide='none') +
    guides(color=guide_legend(ncol=4),
           shape=guide_legend(ncol=4)) +
    labs(x=x[[2]],y=x[[3]],title=x[[1]],fill='',color='',size=''))

@@

```{r FigS3D, fig.width=3.25 * fig_scale, fig.height=1.25 * fig_scale}
read.xlsx('data/source_data.xlsx', sheet='HTGTS homology') %>%
  dplyr::mutate(Genotype=factor(Genotype, levels=c('WT','Atm','Lig4','Trp53bp1')),
                celltype=gsub('spleen','Spleen',celltype),
                n_homology=as.numeric(n_homology)) %>%
  dplyr::filter(xor(isotype=='Sa', celltype=='Spleen')) %>%
  dplyr::group_by( celltype, Genotype, isotype, n_homology) %>%
  dplyr::summarise(mean_frac=mean(frac),
                   se_frac=sd(frac)/sqrt(n())) %>%
  ggplot(aes(x=n_homology, y=mean_frac, color=Genotype)) +
  geom_line(aes(group=Genotype)) +
  geom_errorbar(aes(ymin=mean_frac-se_frac,ymax=mean_frac+se_frac),width=0) +
  ggh4x::facet_nested(. ~ celltype  + isotype , scales='free', space='free_x', nest_line = TRUE, switch='y') +
  my_theme() +
  scale_color_manual(values=genotype_colors) +
  labs(color='', y='Fraction junctions') +
  coord_cartesian(clip='off') 
```

@@

```{r Fig3B, fig.width=3.5 * fig_scale, fig.height=1.75 * fig_scale}
CH12.data <- read.xlsx('data/source_data.xlsx',sheet='CH12 data') %>%
  tibble::column_to_rownames('Sample')
CH12.data.noSg <- read.xlsx('data/source_data.xlsx',sheet='CH12 data (no Sg)') %>%
  tibble::column_to_rownames('Sample')
mouse.data.noSg <- read.xlsx('data/source_data.xlsx',sheet='mouse tissues (no Sg)')
